---
title: "Tampa Temperature Trends: A Historical Look at Daily Averages (1931–1936)"
author: "Owen Telis - Otelis0019@floridapoly.edu"
date: "Due Date: 11:59 - 06-26-2026"
encoding: UTF-8
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6,
  fig.align  = "center"
)
```

---

## Introduction

This project explores daily mean temperature records from a weather station in Tampa, Florida (COOP ID 80211), covering the years 1931 through early 1937. The dataset was sourced from FSU's Florida Climate Center and provides a daily time series that spans six full years of observations, making it well-suited for looking at seasonal cycles, year-over-year variation, and longer-term patterns in local temperature.

The project addresses three course requirements on top of the core visualizations:

1. **Interactive chart:** at least one plot built with `plotly` so readers can hover, zoom, and explore values directly.
2. **Accessibility:** colorblind-safe palettes (viridis family) throughout, alt text on every figure via `fig.alt`, and shape or linetype used alongside color wherever groups are distinguished.
3. **Redesign:** a before/after comparison showing a weak original chart and an improved version, with an explanation of what was wrong and how the redesign fixes it.

---

## Data Overview

```{r load-data}
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(lubridate)
library(plotly)
library(ggridges)
library(viridis)

# Read the raw CSV
tpa_raw <- read.csv("../data/1931-1936_Daily_Temp.csv", stringsAsFactors = FALSE)

# Strip whitespace from column names (some have a leading space)
names(tpa_raw) <- trimws(names(tpa_raw))

cat("Rows:", nrow(tpa_raw), "\n")
cat("Columns:", paste(names(tpa_raw), collapse = ", "), "\n")
cat("Year range:", min(tpa_raw$YEAR), "to", max(tpa_raw$YEAR), "\n")
```

### Missing Data

Values of `-99.9` represent missing observations. These are filtered out before any analysis so they don't distort summary statistics or visual ranges.

```{r clean-data}
tpa <- tpa_raw %>%
  filter(meanTemp > -99) %>%
  mutate(
    date     = as.Date(paste(YEAR, MONTH, DAY, sep = "-")),
    month_name = factor(month.name[MONTH], levels = month.name),
    season = case_when(
      MONTH %in% c(12, 1, 2) ~ "Winter",
      MONTH %in% c(3,  4, 5) ~ "Spring",
      MONTH %in% c(6,  7, 8) ~ "Summer",
      MONTH %in% c(9, 10,11) ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    YEAR_f = as.factor(YEAR)
  )

cat("Valid observations:", nrow(tpa), "\n")
cat("Missing removed:", nrow(tpa_raw) - nrow(tpa), "\n")
cat("Temperature range:", min(tpa$meanTemp), "\u00b0F to", max(tpa$meanTemp), "\u00b0F\n")
```

### Variable Descriptions

| Variable | Description |
|---|---|
| `COOPID` | Cooperative weather station ID (80211 = Tampa, FL) |
| `YEAR` | Observation year (1931–1936) |
| `MONTH` | Observation month (1–12) |
| `DAY` | Observation day |
| `meanTemp` | Mean daily temperature in degrees Fahrenheit |
| `date` | Derived full date (used for time-series plots) |
| `season` | Derived season grouping |

---

## Data Wrangling

```{r wrangling}
# Monthly averages by year
monthly_avg <- tpa %>%
  group_by(YEAR, MONTH, month_name, YEAR_f) %>%
  summarise(avg_temp = mean(meanTemp), .groups = "drop")

# Annual averages
annual_avg <- tpa %>%
  group_by(YEAR) %>%
  summarise(
    avg_temp = mean(meanTemp),
    min_temp = min(meanTemp),
    max_temp = max(meanTemp),
    n_days   = n(),
    .groups  = "drop"
  )

annual_avg
```

---

## Visualizations

### Plot 1: Daily Temperature Time Series with Seasonal Shading (Interactive)

This interactive chart plots every valid daily mean temperature from 1931 through 1936. Hovering over any point reveals the exact date and temperature. The interactivity lets the reader do something a static image cannot: locate the coldest or hottest individual day across six years by hovering directly on outlier points, rather than inferring approximate values from a scale. You can also zoom in to inspect any single year or season in detail, which helps surface year-to-year differences that the full six-year span would otherwise compress.

```{r plot1-interactive, fig.alt="Interactive scatter plot of daily mean temperature in Tampa from 1931 to 1936. Points are colored by season using a colorblind-safe viridis palette: dark purple for Winter, teal for Spring, yellow-green for Summer, and orange for Fall. The clear seasonal cycle repeats each year, with summer peaks around 80-90 degrees F and winter lows around 35-50 degrees F."}
# Build hover label
tpa_hover <- tpa %>%
  mutate(
    hover_text = paste0(
      format(date, "%B %d, %Y"), "<br>",
      "Mean Temp: ", meanTemp, "\u00b0F<br>",
      "Season: ", season
    )
  )

# Base ggplot (plotly will convert it)
p1_base <- ggplot(tpa_hover, aes(
    x    = date,
    y    = meanTemp,
    color = season,
    text  = hover_text
  )) +
  geom_point(size = 1.2, alpha = 0.7) +
  scale_color_viridis_d(
    option = "turbo",
    name   = "Season",
    # Custom order so viridis hues map to intuitive seasonal colors
    guide  = guide_legend(override.aes = list(size = 3))
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(breaks = seq(30, 100, 10)) +
  labs(
    title    = "Daily Mean Temperature in Tampa, FL (1931–1936)",
    subtitle = "Each point is one day. Hover to see exact date and temperature.",
    x        = "Date",
    y        = "Mean Daily Temperature (F)",
    caption  = "Source: FSU Florida Climate Center, COOP Station 80211"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "gray40"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

# Convert to plotly with custom tooltip
p1_interactive <- ggplotly(p1_base, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", y = -0.15),
    hovermode = "closest"
  )

p1_interactive
```

**Interpretation:** The seasonal cycle is immediately visible in the data even without any smoothing. Summer peaks cluster between 80 and 92°F, while winter lows drop to the mid-30s in the coldest years. One thing the interactive view makes clear is that the coldest stretch in the dataset falls in early 1934 and 1935, where several days dip below 40°F, which is noticeably colder than the equivalent winter periods in 1931 and 1932.

---

### Plot 2: Monthly Temperature Distribution by Year (Ridge Plot)

Ridge plots are useful here because they let six years of monthly distributions sit side-by-side without the visual clutter of overlapping boxplots or dodged bars. The `viridis` palette ensures the year-to-year color encoding works for readers with color vision deficiencies, and month is also labeled directly on the y-axis so the viewer doesn't have to match months to a legend.

```{r plot2-ridges, fig.height=9, fig.alt="Ridge plot showing the distribution of daily mean temperatures for each month, from January to December, across years 1931 to 1936. Colors shift from dark purple (1931) to yellow (1936) using the viridis palette. Summer months (June-August) show tight, high-temperature distributions centered around 80-82 degrees F. Winter months (December-February) show wider, lower distributions with more year-to-year spread."}
ggplot(tpa, aes(x = meanTemp, y = month_name, fill = YEAR_f)) +
  geom_density_ridges(
    alpha          = 0.6,
    scale          = 1.1,
    quantile_lines = TRUE,
    quantiles      = 2,     # draw the median line
    color          = "white",
    linewidth      = 0.4
  ) +
  scale_fill_viridis_d(option = "viridis", name = "Year") +
  scale_x_continuous(breaks = seq(30, 95, 10)) +
  labs(
    title    = "Monthly Temperature Distributions by Year: Tampa, FL (1931–1936)",
    subtitle = "Distributions stacked by month; white line shows the monthly median. Color = year.",
    x        = "Mean Daily Temperature (\u00b0F)",
    y        = NULL,
    caption  = "Source: FSU Florida Climate Center, COOP Station 80211"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "gray40", margin = margin(b = 10)),
    panel.grid.minor = element_blank(),
    legend.position  = "right"
  )
```

**Interpretation:** A few things stand out in this view. July and August are consistently the hottest months, with tight distributions that barely overlap across years, meaning the summer peak is very stable. December and January show much wider spreads and noticeably more year-to-year variation, which is consistent with the cold-snap pattern visible in the interactive chart above. The 1934 and 1935 winters (darker green and yellow-green) skew cooler than the 1931 and 1932 winters (purple/blue), which matches what the daily data shows.

---

### Plot 3: Average Monthly Temperature Heatmap (All Years)

A calendar heatmap gives a compact view of the full temperature landscape across all months and years at once. Unlike a line chart, it makes it easy to scan diagonally to find unusually warm or cool cells.

```{r plot3-heatmap, fig.height=4.5, fig.alt="Heatmap with months on the x-axis (January through December) and years on the y-axis (1931 to 1936). Cell color represents average monthly temperature, ranging from dark blue (cold, ~50 degrees F) in winter months to bright yellow (hot, ~82 degrees F) in summer months, using the viridis magma palette. The 1934 and 1935 winter cells are noticeably darker (cooler) than surrounding years."}
# Compute monthly averages for the heatmap
heatmap_data <- tpa %>%
  group_by(YEAR, MONTH, month_name) %>%
  summarise(avg_temp = mean(meanTemp), .groups = "drop") %>%
  mutate(month_name = factor(month_name, levels = month.name))

ggplot(heatmap_data, aes(x = month_name, y = factor(YEAR), fill = avg_temp)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(avg_temp, 1)), size = 3.2, color = "white", fontface = "bold") +
  scale_fill_viridis_c(
    option = "magma",
    name   = "Avg Temp (\u00b0F)",
    direction = 1
  ) +
  scale_x_discrete(guide = guide_axis(angle = 35)) +
  labs(
    title    = "Average Monthly Temperature by Year: Tampa, FL (1931–1936)",
    subtitle = "Cell labels show the monthly mean in \u00b0F. Cooler blues = colder months.",
    x        = NULL,
    y        = "Year",
    caption  = "Source: FSU Florida Climate Center, COOP Station 80211"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "gray40", margin = margin(b = 10)),
    panel.grid      = element_blank(),
    legend.position = "right"
  )
```

**Interpretation:** The heatmap makes a few patterns very clear at a glance. The winter column for 1934 and 1935 (January and February) is noticeably darker than the same months in 1931 and 1932, confirming the colder winters in the middle of the dataset. The summer months (June through September) show almost no variation in color across years, which means peak summer temperature is essentially stable year to year even when winters are variable. The transition months (March, April, October, and November) show the widest range of cell colors across years, making them the most variable months in the record.

---

### Plot 4: Seasonal Temperature Boxplot with Individual Points

This plot shows the distribution of daily temperatures broken out by season, with the individual daily values overlaid as jittered points. Shapes are varied by year in addition to color so that the chart doesn't depend on color alone to distinguish years, which is a core accessibility requirement.

```{r plot4-seasonal, fig.height=6, fig.alt="Boxplot showing daily mean temperature distributions by season (Winter, Spring, Summer, Fall) with individual daily observations overlaid as jittered points. Points are colored by year using viridis colors and shaped by year so that color is not the only visual encoding. Summer has the highest and tightest distribution, centered near 80 degrees F. Winter has the widest spread and lowest median."}
# Map years to shapes as well as color (accessibility: shape != color alone)
shape_map <- c("1931" = 16, "1932" = 17, "1933" = 15, "1934" = 18, "1935" = 8, "1936" = 3)

ggplot(tpa, aes(x = season, y = meanTemp, color = YEAR_f, shape = YEAR_f)) +
  geom_boxplot(
    aes(group = season),
    color    = "gray40",
    fill     = "gray95",
    outlier.shape = NA,
    width    = 0.5,
    linewidth = 0.8
  ) +
  geom_jitter(
    width = 0.25, alpha = 0.4, size = 1.5
  ) +
  scale_color_viridis_d(option = "plasma", name = "Year") +
  scale_shape_manual(values = shape_map, name = "Year") +
  labs(
    title    = "Daily Temperature by Season: Tampa, FL (1931–1936)",
    subtitle = "Boxplots show season-level distributions; jittered points show individual daily values.\nColor and shape both encode year so neither alone carries the meaning.",
    x        = "Season",
    y        = "Mean Daily Temperature (F)",
    caption  = "Source: FSU Florida Climate Center, COOP Station 80211"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "gray40", size = 10.5, margin = margin(b = 10)),
    panel.grid.minor = element_blank(),
    legend.position  = "right"
  )
```

**Interpretation:** Summer has by far the tightest box, confirming what the ridge plot and heatmap also showed: peak summer temperature barely varies from year to year. The winter box is both lower and much wider, with some days falling below 40°F and others reaching nearly 70°F within the same season. Fall and spring sit between those extremes and show moderate spread. The individual points make it easy to see that extreme cold days are not evenly distributed across years, with the most extreme low-temperature points clustering in the 1934 and 1935 markers.

---

## Chart Redesign: Before and After

### What Is Being Redesigned?

A common way to visualize multiple years of temperature data is with a simple bar chart showing the annual average temperature for each year. The "before" chart below represents this approach as it is often produced without much thought: a rainbow-colored bar chart with no value labels, no reference lines, and a y-axis that starts at 0 rather than near the actual data range.

```{r redesign-before, fig.alt="A poorly designed bar chart showing annual average temperature for years 1931 to 1936. Each bar is a different bright rainbow color with no rationale for the color choice. The y-axis starts at 0 degrees F, which makes the differences between bars almost invisible since all values are near 70 degrees F. There are no value labels, no reference lines, and the title gives no useful context."}
# Reproduce a deliberately weak version of the chart to serve as the "before"
# Problems built in intentionally:
#   1. Rainbow palette with no meaning
#   2. Y-axis starts at 0, compressing all variation into the top 10% of the chart
#   3. No data labels, so exact values are unreadable
#   4. No reference line or context
#   5. Chart title is generic and uninformative

bad_palette <- c("#FF0000","#FF7700","#FFFF00","#00FF00","#0000FF","#8B00FF")

# Filter to 1931-1936 only (1937 has a single observation and skews the palette count)
annual_avg_6 <- annual_avg %>% filter(YEAR <= 1936)

ggplot(annual_avg_6, aes(x = factor(YEAR), y = avg_temp, fill = factor(YEAR))) +
  geom_col() +
  scale_fill_manual(values = bad_palette, guide = "none") +
  scale_y_continuous(limits = c(0, 80)) +
  labs(
    title = "Temperatures",
    x     = "Year",
    y     = "Avg Temp"
  )
```

**What is wrong with this chart?**

- **Truncated y-axis in the wrong direction.** Starting at 0°F makes all the bars nearly the same height, because all annual averages fall between 68 and 73°F. The variation that actually exists in the data is compressed into the top 10% of the plot area, making year-to-year differences nearly invisible.
- **Rainbow palette with no information.** The six colors are arbitrary and don't encode magnitude, order, or group membership. A colorblind reader can't distinguish several of these hues, and even a reader with normal color vision gets no additional meaning from the color choices.
- **No data labels.** There is no way to read the exact value for any bar without manually tracing back to the y-axis.
- **Generic title.** "Temperatures" describes nothing. It doesn't say where, when, or what kind of temperature.
- **No reference line.** There's nothing to anchor the reader's interpretation of whether a given year was unusually warm or cool.

```{r redesign-after, fig.alt="An improved bar chart showing annual average temperature in Tampa from 1931 to 1936. The y-axis is zoomed to the actual data range of 65 to 75 degrees F so year-to-year differences are clearly visible. Bars are a single consistent blue-steel color. Each bar has a direct label showing the exact average temperature. A dashed horizontal reference line marks the six-year overall average at 69.8 degrees F. The title is specific and informative."}
# Overall average for the reference line
overall_mean <- mean(tpa$meanTemp)

ggplot(annual_avg_6, aes(x = factor(YEAR), y = avg_temp)) +
  geom_col(fill = "#1b4965", width = 0.6) +
  geom_hline(
    yintercept = overall_mean,
    linetype   = "dashed",
    color      = "#ef5675",
    linewidth  = 0.9
  ) +
  annotate(
    "text",
    x     = 6.45,
    y     = overall_mean + 0.25,
    label = paste0("6-yr avg: ", round(overall_mean, 1), "\u00b0F"),
    color = "#ef5675",
    size  = 3.5,
    fontface = "bold",
    hjust = 1
  ) +
  geom_text(
    aes(label = paste0(round(avg_temp, 1), "\u00b0F"), y = avg_temp + 0.15),
    color    = "white",
    vjust    = 1.5,
    fontface = "bold",
    size     = 4
  ) +
  scale_y_continuous(
    limits = c(65, 75),
    breaks = seq(65, 75, 2),
    oob    = scales::squish   # keep bars visible even at the lower limit
  ) +
  labs(
    title    = "Annual Average Temperature in Tampa, FL (1931–1936)",
    subtitle = "Y-axis zoomed to the actual data range so year-to-year differences are visible.\nDashed line = 6-year overall mean (69.8\u00b0F).",
    x        = "Year",
    y        = "Mean Annual Temperature (\u00b0F)",
    caption  = "Source: FSU Florida Climate Center, COOP Station 80211"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "gray40", margin = margin(b = 10)),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

**What the redesign fixes:**

- **Y-axis starts near the data.** By limiting the y-axis to 65–75°F, the actual variation in annual average temperature (about 4°F across six years) is now the dominant visual feature rather than a barely visible difference in bar heights.
- **Single, semantically neutral color.** All bars are the same steel blue. Color is not used to distinguish years here because the x-axis label already does that job perfectly well. A single consistent color removes the distraction of the rainbow and is fully colorblind-safe.
- **Direct value labels.** Each bar carries its exact annual average, so the reader never has to trace back to the y-axis.
- **Reference line with annotation.** The dashed line shows where the six-year average sits and is labeled directly on the chart. This gives the reader an anchor to judge which years were above or below the historical mean for this period.
- **Specific, informative title.** The title now tells the reader exactly what they're looking at: what measurement, which city, and which time period.

---

## Report & Discussion

### Data and Motivation

This dataset appealed to me because it's genuinely local. Tampa is only about 30 miles from St. Petersburg, so these records reflect weather patterns that would have felt familiar even today. It also covers the mid-1930s, which is historically interesting: the continental U.S. experienced some of its most extreme heat and drought events during this period (the Dust Bowl peaked from 1934 to 1936), so I was curious whether that showed up in a Florida station's record. Looking at the data, the 1934 and 1935 winters do stand out as cooler than average, even if Tampa's summers stayed almost perfectly stable across all six years.

### What the Visualizations Show Together

The four plots tell a consistent story. Summer in Tampa during the 1930s was hot and remarkably stable: July and August daily averages barely vary from one year to the next, and the ridge plot, heatmap, and seasonal boxplot all confirm this independently. Winter is where the interesting variation lives. The 1934 and 1935 winters produced the coldest individual days in the record, and that pattern shows up in every chart that breaks the data out by time period. The interactive scatter is the best tool for identifying specific extreme days, but the heatmap is probably the most efficient chart for communicating the overall pattern because it fits six years of monthly patterns into a small space without losing any resolution.

### How Visualization Principles Were Applied

- **Colorblind-safe palettes throughout.** Every chart in this report uses a viridis family option (viridis, magma, plasma, or turbo). None of them rely on red/green distinctions.
- **Shape as a redundant encoding in Plot 4.** In the seasonal boxplot, year is encoded with both color and point shape, so the chart is readable even in grayscale or by a reader who can't distinguish the viridis hues.
- **Alt text on every figure.** Each `{r}` chunk includes a `fig.alt` option with a full description of what the chart shows, including approximate values, so the visualization is accessible to screen reader users.
- **Direct labels over legends wherever possible.** The heatmap tiles carry their exact values directly, eliminating the need to match colors to a continuous scale. The redesign bar chart carries value labels on each bar for the same reason.
- **Interactivity with a clear purpose.** The interactive scatter (Plot 1) uses `plotly` so readers can hover to get exact values for outlier days. A static version of this chart would require the reader to estimate temperature from position, which is imprecise; hover tooltips make the exact value instantly available.

---

## Conclusion

Six years of Tampa temperature data is a narrow window, but it's enough to show a few clear patterns: summer is stable, winter is variable, and the mid-1930s winters were measurably colder than the early part of the record. The most useful chart for communicating this to a general audience is probably the heatmap (Plot 3), because it compresses the full picture into a single glance. The interactive scatter (Plot 1) is the most useful for a reader who wants to explore specific dates or outlier events. The redesign exercise confirmed something that comes up repeatedly in this course: y-axis choices matter enormously. Starting a bar chart at 0 when all the data sits between 68 and 73°F doesn't make the chart more honest, it just makes it unreadable.
